---
title: "**Dependent *t*-Tests**"
subtitle: "**STA4173: Biostatistics** <br> Spring 2025"
execute:
  echo: true
  warning: false
  message: false
  error: true
format: 
  revealjs:
    code-overflow: wrap
    df-print: paged
    embed-resources: true
    slide-number: false
    width: 1600
    height: 900
    html-math-method: katex
    theme:
      - default
      - sp25.scss
editor: source
---

## **Introduction**

- In the last lecture, we reviewed statistical inference on two independent means.
    
    - CI for $\mu_1-\mu_2$
    - Hypothesis test for $\mu_1-\mu_2$ (two-sample *t*-test)

- Today, we will focus on drawing *conclusions* about two dependent means.

    - CI for $\mu_d = \mu_1-\mu_2$
    - Hypothesis test for $\mu_d = \mu_1-\mu_2$ (paired *t*-test)

## **Independent vs. Dependent Data**

::: {.callout-note}
## Independent data 
An individual selected for one sample does not dictate which individual is to be in a second sample.

In the data, there is not a way to link the individuals in the sample.
:::

::: {.callout-note}
## Dependent data 
An individual selected to be in one sample is used to determine the individual in the second sample.

In the data, there is a way to link the individuals in the sample.
:::

- Examples:
    - Two sections of STA4173
    - Project grades in one section of STA4173
    - Male and female penguins
    - Prices online vs. in store at Target

## **Estimating $\mathbf{\boldsymbol \mu_d = \boldsymbol \mu_1-\boldsymbol \mu_2}$**

-   We are now interested in comparing two **dependent** groups.

-   We assume that the two groups come from the same population and are going to examine the difference,

$$
d = y_{i, 1} - y_{i, 2}
$$

-   After drawing samples, we have the following,
    -   $\bar{d}$ estimates $\mu_d$,
    -   $s^2_d$ estimates $\sigma^2_d$, and
    -   $n$ is the sample size.

## **Confidence Interval for $\mathbf{\boldsymbol \mu_d}$**

::: {.callout-note}
## $\mathbf{(1-\boldsymbol\alpha)100\%}$ confidence interval for $\mathbf{\boldsymbol\mu_d}$

$$ \bar{d} \pm t_{\alpha/2} \frac{s_d}{\sqrt{n}} $$

-   where $t_{\alpha/2}$ has $n-1$ degrees of freedom.
-   To construct this interval, we require either:
    -   the differences to be normally distributed or
    -   the sample size is sufficiently large ($n \ge 30$)

:::

- `R` syntax:

```{r}
#| eval: false
t.test(dataset_name$variable1_name,
       dataset_name$variable2_name, 
       paired = TRUE, 
       conf.level = confidence_level)
```

## **Confidence Interval for $\mathbf{\boldsymbol \mu_d}$: Example**

-   Insurance adjusters are concerned about the high estimates they are receiving for auto repairs from garage I compared to garage II.
-   15 cars were taken to both garages for separate estimates of repair costs.

```{r}
library(tidyverse)
garage <- tibble(g1 = c(17.6, 20.2, 19.5, 11.3, 13.0, 
                        16.3, 15.3, 16.2, 12.2, 14.8,
                        21.3, 22.1, 16.9, 17.6, 18.4), 
                 g2 = c(17.3, 19.1, 18.4, 11.5, 12.7, 
                        15.8, 14.9, 15.3, 12.0, 14.2, 
                        21.0, 21.0, 16.1, 16.7, 17.5))
```

-   Construct the 95% confidence interval for the average difference between the two garages.

- Remember the `R` syntax:

```{r}
#| eval: false
t.test(dataset_name$variable1_name,
       dataset_name$variable2_name, 
       paired = TRUE, 
       conf.level = confidence_level)
```


## **Confidence Interval for $\mathbf{\boldsymbol \mu_d}$: Example**

```{r}
t.test(garage$g1, 
       garage$g2, 
       paired = TRUE, 
       conf.level = 0.95)
```

-   The 95% CI for $\mu_d$, where $d = x_{\text{I}} - x_{\text{II}}$ is (0.39, 0.83).

-   From the problem statement:
    - *Insurance adjusters are concerned about the high estimates they are receiving for auto repairs from garage I compared to garage II.*
-   Can we say that estimates from garage I are higher than those from garage II?

## **Paired *t*-Test**

::: {.callout-note}
## Hypothesis Test for Two Dependent Means
Hypotheses

- $H_0:$ $\mu_d = \mu_0$ $|$ $H_0:$ $\mu_d \le \mu_0$ | $H_0:$ $\mu_d \ge \mu_0$
- $H_1:$ $\mu_d \ne \mu_0$ | $H_0:$ $\mu_d > \mu_0$ | $H_1:$ $\mu_d < \mu_0$ 

Test Statistic
$$
t_0 = \frac{\bar{d}-\mu_0}{\frac{s_d}{\sqrt{n}}}
$$

*P*-Value

- $p = 2 P[t \ge |t_0|]$ | $p = P[t \ge |t_0|]$ | $p = P[t \le |t_0|]$

Rejection Region

- Reject $H_0$ if $p < \alpha$.

Conclusion/Interpretation

- [Reject or fail to reject] $H_0$.

- There [is or is not] sufficient evidence to suggest [alternative hypothesis in words].

:::

## **Paired *t*-Test**

- R syntax:

```{r}
#| eval: false
t.test(dataset_name$variable1_name, 
       dataset_name$variable2_name, 
       paired = TRUE, 
       mu = hypothesized_difference,
       alternative = "alternative")
```

- Important!!
    - We are estimating $\mu_1 - \mu_2$, but R is going to subtract in the order we state in the `t.test()` function.
        - e.g., if we want to examine $\mu_d  = \mu_{\text{freshman}} - \mu_{\text{sophomore}}$:

```{r}
#| eval: false
t.test(students$freshman_gpa, 
       students$sophomore_gpa, 
       paired = TRUE, 
       mu = 0,
       alternative = "alternative")
```

```{r}
#| eval: false
t.test(students$sophomore_gpa, 
       students$freshman_gpa, 
       paired = TRUE, 
       mu = 0,
       alternative = "alternative")
```

## **Paired *t*-Test: Example**

-   Let's now formally determine if garage I's estimates are higher than garage II's. Test at the $\alpha=0.05$ level.

- Recall the data,

```{r}
#| eval: false
garage <- tibble(g1 = c(17.6, 20.2, 19.5, 11.3, 13.0, 
                        16.3, 15.3, 16.2, 12.2, 14.8,
                        21.3, 22.1, 16.9, 17.6, 18.4), 
                 g2 = c(17.3, 19.1, 18.4, 11.5, 12.7, 
                        15.8, 14.9, 15.3, 12.0, 14.2, 
                        21.0, 21.0, 16.1, 16.7, 17.5))
```

- and the `R` syntax:

```{r}
#| eval: false
t.test(dataset_name$variable1_name, 
       dataset_name$variable2_name, 
       paired = TRUE, 
       mu = hypothesized_difference,
       alternative = "alternative")
```

## **Paired *t*-Test: Example**

```{r}
t.test(garage$g1, 
       garage$g2,
       paired = TRUE,
       mu = 0,
       alternative = "greater")
```

- Are the estimates from garage I significantly higher than those from garage II?

## **Paired *t*-Test: Example**

**Hypotheses**

-   $H_0: \ \mu_{\text{I}} \le \mu_{\text{II}}$ OR $\mu_{d} \le 0$, where $\mu_d = \mu_{\text{I}} - \mu_{\text{II}}$
-   $H_1: \ \mu_{\text{I}} > \mu_{\text{II}}$ OR $\mu_{d} > 0$

**Test Statistic and *p*-Value**

- $t_0 = 6.023$
- $p < 0.001$

**Rejection Region**

- Reject $H_0$ if $p < \alpha$; $\alpha = 0.05$.

**Conclusion/Interpretation**

- Reject $H_0$.
- There is sufficient evidence to suggest the estimates at garage I are higher than that of garage II.

## **Wrap Up**

- Today we reviewed the dependent *t*-test.
    - Confidence intervals
    - Hypothesis testing
    
- Get to know you quiz - complete with RStudio - was due <u>Monday or Tuesday</u>.

- <u>Campus is closed on Monday!</u>
    - Tuesday: *optional* session for Assignment 1 help.
    - TR section: 2:30-3:45 pm, 4/310.